A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-21 paper.tex
Address some comments

\section*{Introduction}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
% TODO say this better.
From a high-level, ARGs describe the sequence of correlated genealogical
trees along the genome, induced by the different paths of
genetic inheritance caused by recombination.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local trees
is somewhat subtle (see section XXX), and it is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, $\pi_u = -1$.
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

\section*{Event ARGs}\label{eARG}
Genome ARGs are concerned with
describing the details of genetic inheritance
between specific pairs of genomes.
In contrast, ARGs are classically seem defined
by the evolutionary \emph{events} in the history of a sample.
The difference between gARGs and this prevalent view of
what defines an ARG is subtle and the distinctions sometimes porous.
It is therefore useful to first survey some representative
definitions of
an ARG from the literature, to allow us to capture the common features
of these definitions
(see Appendix XXX for a full review).

>>>>

\section*{Introduction}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
% TODO say this better.
From a high-level, an ARG describes the different paths of genetic inheritance 
caused by recombination, and encodes a sequence of correlated genealogical 
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,fan2022genealogical,hejase2022deep,zhang2023biobank,
nowbandegani2023extremely}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.

@@

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local trees
is somewhat subtle (see section XXX), and it is useful to have
a concrete definition and explicit process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this reachability stuff is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, $\pi_u = -1$ (by convention).
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an find edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. We then set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

@@

\section*{Event ARGs}\label{eARG}
Genome ARGs are composed by 
describing the details of genetic inheritance
between specific pairs of genomes.
In contrast, ARGs are classically seem defined
by the evolutionary \emph{events} in the history of a sample.
The difference between gARGs and this prevalent view of
what defines an ARG is subtle and the distinctions sometimes porous.
It is therefore useful to first survey some representative
definitions of
an ARG from the literature, to allow us to capture the common features
of these definitions
(see Appendix XXX for a full review).
